        ---
title: "SDS 322E: Project 1 Report"
output: html_document
---

## Names and EIDs of Group Members
Siddhi Bansasl sb63855
Ananth Kothuri ask3343
Maehek Bhatt mb66256


## Load the Data

```{r setup, include=FALSE}
## Do not modify this code chunk
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```



```{r, cache=FALSE, message=FALSE}
## This chunk may take a long time to run. Set 'cache=TRUE' in the chunk options
## to make it run faster in subsequent knittings.

championships <- read_tsv("WCA_export_championships.tsv.bz2")
competitions <- read_tsv("WCA_export_Competitions.tsv.bz2")
continents <- read_tsv("WCA_export_Continents.tsv.bz2")
countries <- read_tsv("WCA_export_Countries.tsv.bz2")
eligible_country_iso2s_for_championship <- read_tsv("WCA_export_eligible_country_iso2s_for_championship.tsv.bz2")
events <- read_tsv("WCA_export_Events.tsv.bz2")
formats <- read_tsv("WCA_export_Formats.tsv.bz2")
persons <- read_tsv("WCA_export_Persons.tsv.bz2")
ranksaverage <- read_tsv("WCA_export_RanksAverage_333.tsv.bz2")
rankssingle <- read_tsv("WCA_export_RanksSingle_333.tsv.bz2")
results <- read_tsv("WCA_export_Results_333.tsv.bz2")
roundtypes <- read_tsv("WCA_export_RoundTypes.tsv.bz2")
scrambles <- read_tsv("WCA_export_Scrambles.tsv.bz2")
```


## Questions

### Active Speed Cubers

How many active (3x3x3) speedcubers are there registered with the WCA? For this question an *active speeedcuber* is defined as any person registered in the WCA who has competed in at least two competitions in the years 2022--2024. 


```{r}
# look at the results table, filter by 333 event ID, join with competitions (using competitionId in results and id in competitions), filter by year, group by personId, count the number of competitions, filter by >= 2
# then count the number of unique personIds
num_active_speedcubers <- results |>
  filter(eventId == "333") |>
  inner_join(competitions, by = c("competitionId" = "id")) |>
  filter(year >= 2022 & year <= 2024) |>
  group_by(personId) |>
  summarise(n_competitions = n()) |>
  filter(n_competitions >= 2) |>
  summarise(n_active_speedcubers = n_distinct(personId))

print(num_active_speedcubers)

```

**There are 50,439 active speedcubers registered with the WCA.**

### World Records

This question has two parts:

1. Who holds the current world record single? 

```{r}
## Filter rankssingle by worldRank == 1
curr_world_record <- rankssingle |>
  filter(worldRank == 1) |>
  inner_join(persons, by = c("personId" = "id")) |>
  select(name, personId)
print(curr_world_record)

```

**Max Park holds the current world record single.**

On what date was this record set?


```{r}
# passed their id to find the date of the record
world_record_date <- results |>
    filter(personId == "2012PARK03", pos == 1, eventId == "333") |>
    arrange(best) |>
    slice(1) |>
    inner_join(competitions, by = c("competitionId" = "id")) |>
    select(year, month, day)

print(world_record_date)
```

**The world record was set on June 11, 2023.**



2. Who *previously* held the world record single? On what date was this previous record set?

```{r}
# found the previous world record holder
curr_world_record <- rankssingle |>
  filter(worldRank == 2) |>
  inner_join(persons, by = c("personId" = "id")) |>
  select(name, personId)
print(curr_world_record)

# passing their id to find the date of world record
world_record_date <- results |>
    filter(personId == "2017GARR05", pos == 1, eventId == "333") |>
    arrange(best) |>
    slice(1) |>
    inner_join(competitions, by = c("competitionId" = "id")) |>
    select(year, month, day)

print(world_record_date)
```

**The previous world record holder was Luke Garret who set the record on February 25, 2023.**


### Regional Rankings

This question has two parts:

1. Amongst all speedcubers, who is the top ranked male speedcuber (for best single solve) in Australia?

```{r}
# filtered persons by gender and country
males <- persons |>
    filter(countryId == "Australia", gender == "m") |>
    inner_join(rankssingle, by = c("id" = "personId")) |>
    filter(countryRank == 1) |>
    select(id, name)

print(males)
```

**The top ranked male speedcuber in Australia is Jode Brewster.**


2. Amongst all speedcubers, who is the top ranked female speedcuber (for best single solve time) in Europe?

```{r}
# joined countries with persons to filter by continent and gender
females <- persons |>
    inner_join(countries, by = c("countryId" = "id")) |>
    filter(gender == "f", continentId == "_Europe") |>
    inner_join(rankssingle, by = c("id" = "personId")) |>
    arrange(best) |>
    slice(1) |>
    select(id, name.x)

print(females)

```

**The top ranked female speedcuber in Europe is Magdalena Pabisz.**



### Time Until Sub-5

Having a time below 5 seconds is considered an elite achievement and most speedcubers have to complete a large number of solves before they can obtain a sub-5 second solve. 

**NOTE**: Each round of a competition has 5 solves that should be considered separately when counting the number of solves.


1. For the current top 10 speedcubers in the world (as recorded in the RanksSingle table), on average, how many solves did they have to do before achieving a sub-5 second solve?


```{r}
# get top 10 speed cubers
top_cubers <- rankssingle |> 
  arrange(worldRank) |> 
  head(10) |> 
  select(personId)

# pivot table to make it more readable
long_results <- results |> 
  pivot_longer(cols = starts_with("value"), names_to = "solve_number", values_to = "time")

# get all results of only top 10 cubers
top_cuber_results <- long_results |> 
  inner_join(top_cubers, by = "personId")

# join with competitions to get competition date information
top_cuber_results <- top_cuber_results |>
  inner_join(competitions, by = c("competitionId" = "id")) |> 
  select(personId, competitionId, solve_number, time, year, month, day)

#  group results by cuber, and then sort by date
top_cuber_results <- top_cuber_results |> 
  arrange(personId, year, month, day)

# count number of solves it takes for each cuber to get under 5 seconds
sub5_solves <- top_cuber_results |> 
  group_by(personId) |> 
  mutate(solve_index = row_number()) |> 
  filter(time < 500) |> 
  summarise(first_sub5_solve = min(solve_index, na.rm = TRUE))

# get avg number of solves
avg_solves <- mean(sub5_solves$first_sub5_solve, na.rm = TRUE)

# Print result
print(avg_solves)

```

**It took the top 10 speedcubers 132 solves on average to get a sub-5 second solve.**



2. For **one** of the top 10 speedcubers make a plot of their solve times vs. the date of the solve, with date on the x-axis and solve time on the y-axis. 


```{r}
# getting one cuber's results
target_cuber <- top_cuber_results %>% 
  filter(personId== '2012PARK03') %>% 
  mutate(Date = as.Date(paste(year, month, day, sep="-")))

# plot solve times vs. date
ggplot(target_cuber, aes(x = Date, y = time)) + 
  geom_point() + 
  labs(title = "Solve Times Over Time", x = "Date", y = "Solve Time (seconds)")

```


**Write your answer here.**


### Up-and-Coming Speed Cubers

Which speed cubers **not** in the top 10,000 (worldwide for single best time) should we keep an eye on for the near future? 

The idea here is to identify "up-and-coming" speedcubers who are not yet achieving elite times. Come up with a list of **five** speedcubers (provide their names and WCA IDs) that you have identified as "up-and-coming". There is no one way to answer this question and the goal is to provide an analysis of the data that justifies the selection of your five names.

```{r}
# joined the results, rankssingle, and person tables
new_speedcubers <- rankssingle |>
    filter(worldRank > 10000) |>
    inner_join(results, by = c("personId" = "personId")) |>
    inner_join(competitions, by = c("competitionId" = "id")) |>
    group_by(personId) |>
    arrange(personId, year, month, day) |> 
    # calculated the improvement as rate of change from first to last scores
    summarise(
        first_score = first(average),
        last_score = last(average),
        competition_count = n_distinct(competitionId),
        improvement = (first_score - last_score) / competition_count,
        name = first(personName)
    ) |>
    # final filtering
    filter(competition_count > 20) |>
    arrange(desc(improvement)) |>
    slice(1:5)

print(new_speedcubers)
```

**The top up-and-coming speedcubers are 1. 2009NEEL01 - Patitpavan Neel, 2. 2009HEND01 - Paul Hendrickson, 3. 2012IONE01 - Andrei Răzvan Ionel, 4. 2011PALS01 - Shrijan Pal, 5. 2017NOVA05 - João Luiz Melo Novaes. The way I found these speedcubers was to find all competitions for speedcubers under the top 10,000, then chronologically, and finally found their 'rate of improvement' which is calculated as the rate of improvement from their first and last scores. This is then arranged and the top 5 cubers are selected.**



### Region Rivalries

Europe and North America are both regions with strong speedcubers in the WCA. 

Which region has the faster group of speedcubers on average? 

To answer this question, characterize each person using their best *average* score according to their listing in the `ranksaverage` table. In the `persons` table the `countryId` indicates each person's country affiliation. The `countries` table lists the region that each country is in via the `continentId` column (Europe is "_Europe" and North America is "_North America").

Before attempting to answer the question, state what you expect the answer to be below.

**We expect North America to have the faster group of speedcubers than Europe  on average.**


```{r}
## use the ranksaverage table to get the best average score for each person by joining relevant information with persons
## and countries tables, and then get average times per continent

avg_speedcubers <- ranksaverage |>
  inner_join(persons, by = c("personId" = "id")) |>
  inner_join(countries, by = c("countryId" = "id")) |>
  filter(continentId %in% c("_Europe", "_North America")) |>
  group_by(continentId) |>
  summarise(avg_time = mean(best))

print(avg_speedcubers)
```

What do you conclude about speedcubers in Europe vs. North America?

**The average time of speedcubers 3538 in Europe and 3503 in North America. We conclude that North America has the faster group of speedcubers on average.**


### Alternative Explanations

Develop an alternative explanation/hypothesis regarding speedcubers from Europe and North America that is 

1. Consistent with the results you produced in the previous question; but

2. Provides a different interpretation or explanation for what is going on. 

If the results from the previous question were unexpected, make use of systems thinking to develop an alternative hypothesis. If the results were consistent with your expectations, then use skeptical thinking. In either case, you should present an analysis that shows evidence for or against this alternative explanation relative to the conclusion that you made in the previous question.

**Alternative Hypothesis: The average time of North American speedcubers is faster than that of European speedcubers because speedcubing is more popular in North America.**
**If speedcubing is more popular in North America, more people are expected to learn cubing and speedcubing, thus leading to more accessibility and resources as well, which**
**would lead to faster times. This is consistent with the results in the previous question**

**To test this alternative hypothesis, we will count the total number of speedcubers from North America vs. those from Europe.**

```{r}
## count number of competitions for Europe vs. for North America
total_num_speedcubers <- persons |>
    inner_join(countries, by = c("countryId" = "id")) |>
    filter(continentId %in% c("_Europe", "_North America")) |>
    group_by(continentId) |>
    summarise(total_speedcubers = n())

print(total_num_speedcubers)
```

**According to the alternative analysis, there are 58314 North American speedcubers and 53274 European speedcubers. This is consistent with the alternative hypothesis that there are more speedcubers in North American than Europe, and thus**
**could be the reason for the result in the previous question.**




## Discussion

Putting it all together, what did you learn from your data?
• Reflect on the process of conducting this project. What was challenging, what have you
learned from the process itself?

**All of the conclusions that we reached were expected, there was nothing out of the ordinary. We learned more about speedcubing records and**
**how they differ across different demographics.**

**A challenging part of this project was figuring out how to navigate all of the tables to extract relevant data.** 
**Another challenge was making sure that we didn't assume anything about the data when writing code, especially** 
**because there were so many special cases that we had to consider.**

**Completing this project has given us a better understanding of tidyverse functions and gave practice working with**
**complex datasets for real world applications.**

**We worked on the first three questions together, and then split up the next three questions. Maehek did the**
**question about Sub-5 Second Solve, Ananth did the question about Up-and-Coming Speed Cubers, and Siddhi did the**
**question about Region Rivalries. We all worked on the discussion together.**